---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ambiR

<!-- badges: start -->
[![R-CMD-check](https://github.com/NIVA-Denmark/ambiR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/NIVA-Denmark/ambiR/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

This is an R package implementing **AZTI’s Marine Biotic Index**.

## Overview

The package allows the user to calculate both the *AMBI* index and the
multivariate *M-AMBI* from count or abundance data for benthic fauna species.

In addition, this package includes functions to calculate different
versions of the *Danish Quality Index* (*DKI*) (in Danish: *Dansk
Kvalitetsindeks*), a derived benthic index based on AMBI.

-   `AMBI()` - calculates the *AMBI* index.
-   `MAMBI()` - calculates the multivariate *M-AMBI* index.
-   `Hdash()` - calculates *H'*, the Shannon diversity index.
-   `DKI2()` - calculates the salinity-normalised Danish quality index
    *DKI (v2)*.
-   `DKI()` - calculates the Danish quality index *DKI*.

To get started, see `vignette("ambiR")`. For details about running in 
interactive mode, see `vignette("interactive")`.

The AMBI index was developed by Ángel Borja and colleagues at AZTI. For 
background and explanation of the method for calculation of the AMBI index see `vignette("about-AMBI")`. You can also find a link to download the original AMBI
software.

## Installation

You can install the development version of ambiR from [GitHub](https://github.com/niva-denmark/ambiR/) with:

```{r install, eval = FALSE}
# install.packages("devtools")
devtools::install_github("NIVA-Denmark/ambiR")

```

## Examples

This is a basic example using a small dataset.

```{r example}
library(ambiR)

df <- data.frame(station = c("1","1","1","2","2","2"),
                 species = c("Acidostoma neglectum",
                             "Acrocirrus validus",
                             "Capitella nonatoi",
                             "Acteocina bullata",
                             "Austrohelice crassa",
                             "Capitella nonatoi"),
                 count = c(8, 14, 23, 13, 17, 11))

AMBI(df, by = c("station"), format_pct=1)

```

Another example using the supplied `test_data`.

```{r example2}
library(ambiR)

## calling AMBI using the test data set
AMBI(test_data, by=c("station"), var_rep = "replicate", format_pct=1)


```

