---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ambiR

<!-- badges: start -->
[![R-CMD-check](https://github.com/NIVA-Denmark/ambiR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/NIVA-Denmark/ambiR/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

This is an R package implementing **AZTI’s Marine Biotic Index**.

## Overview

The package allows the user to calculate both the *AMBI* index and the
multivariate *M-AMBI* from count data for benthic species.

The original software from AZTI can be downloaded from
<https://ambi.azti.es/>.

In addition, this package includes functions to calculate different
versions of the *Danish Quality Index* (*DKI*) (in Danish: *Dansk
Kvalitetsindeks*), a derived benthic index based on AMBI.

-   `AMBI()` - calculates the *AMBI* index.
-   `MAMBI()` - calculates the multivariate *M-AMBI* index.
-   `Hdash()` - calculates *H'*, the Shannon diversity index.
-   `DKI2()` - calculates the salinity-normalised Danish quality index
    *DKI (v2)*.
-   `DKI()` - calculates the Danish quality index *DKI*.

## Species Groups

Input to the `AMBI()` function is a dataframe of species counts with optional 
grouping variables, e.g. station or replicate IDs. The function matches species
names in the input data with names in the AMBI species list, in order to categorise
the observed species according to the AMBI method. The tool then calculates the
*AMBI* index resulting from the distribution of individuals between the groups.

The AMBI species list gives the groups (I, II, III, IV, V) in which each species 
is classified, as described by Borja et al (2000):

-   *Group I* \
    Species very sensitive to organic enrichment and present under
    unpolluted conditions (initial state). They include the specialist
    carnivores and some deposit- feeding *tubicolous polychaetes*.
-   *Group II* \
    Species indifferent to enrichment, always present in low
    densities with non-significant variations with time (from initial
    state, to slight unbalance). These include suspension feeders, less
    selective carnivores and scavengers.
-   *Group III* \
    Species tolerant to excess organic matter enrichment.
    These species may occur under normal conditions, but their
    populations are stimulated by organic richment (slight unbalance
    situations). They are surface deposit-feeding species, as
    *tubicolous spionids*.
-   *Group IV* \
    Second-order opportunistic species (slight to pronounced
    unbalanced situations). Mainly small sized *polychaetes*: subsurface
    deposit-feeders, such as *cirratulids*.
-   *Group V* \
    First-order opportunistic species (pronounced unbalanced
    situations). These are deposit- feeders, which proliferate in
    reduced sediments.

The list of species and their groups has been updated several times by the 
authors of the AMBI software. The version of the list used here is from 
8th October 2024.


## Installation

You can install the development version of ambiR from [GitHub](https://github.com/niva-denmark/ambiR/) with:

``` r
# install.packages("devtools")
devtools::install_github("NIVA-Denmark/ambiR")

```

## Examples

This is a basic example using a small dataset.

```{r example}
library(ambiR)

df <- data.frame(station = c("1","1","1","2","2","2"),
species = c("Acidostoma neglectum",
             "Acrocirrus validus",
            "Capitella nonatoi",
             "Acteocina bullata",
             "Austrohelice crassa",
             "Capitella nonatoi"),
             count = c(23, 14, 5, 13, 17, 11))

AMBI(df, by = c("station"), format_pct=2)

```

Another example using the supplied `test_data`.

```{r example2}
library(ambiR)

## calling AMBI using the test data set
AMBI(test_data, by=c("station"), var_rep = "replicate", format_pct=2)


```

