---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

This is a brief introduction to calculating AMBI index results.

# Species observation data

Your species counts (or abundance) should be organized in a data frame arranged in *long* format. That is, species names are in one column and species counts in another column. If your data represents several stations and/or if there are replicates, then the data should include columns with this information.

The example dataset included with the package, `test_data`, illustrates how the data should be arranged:

```{r test_data}
library(ambiR)

head(test_data)

```

If your data look like this, you can go directly to [Calculate AMBI]. If not, the following examples show how to reorganize your data.

## data with species names in columns

Here is an example dataframe where there are counts for species in separate columns:

```{r transpose_wide, include=FALSE}
library(dplyr)
library(tidyr)

wide_data_species <- test_data %>%
  pivot_wider(names_from = "species", 
              values_from = "count",
              values_fill = 0)
  
```

```{r data_wide_spec}

head(wide_data_species)
```

To arrange the data in the correct form, use the `tidyr::pivot_longer()` function:

```{r pivot_species}
library(dplyr)
library(tidyr)

# columns 1 and 2 contain station and replicate information
# so, select all columns from 3 to be pivoted 

long_data <- wide_data_species %>%
  pivot_longer(cols = 3:ncol(wide_data_species), 
               names_to = "species",
               values_to = "count")

head(long_data)
```

## data with stations and replicates in columns

Here is an example where each column contains species counts for a station and replicate. The first row of the table contains the station ID *1, 2, ...* and the second row contains the replicate ID *a, b, ...*. The first column of the table contains species names.

```{r transpose_wide_stns, include=FALSE}
library(dplyr)
library(tidyr)

df <- test_data %>%
  pivot_wider(names_from = c("station","replicate"), 
              values_from = "count", values_fill = 0)

stns <- names(df)[2:ncol(df)]

stns <- stns %>% 
  sapply(strsplit,"_") 
reps <- stns  %>%  
  sapply(function(x){x[2]})
stns <- stns  %>%  
  sapply(function(x){x[1]})
n <- 1+length(stns)
df2 <- matrix(nrow=2, ncol=n) %>% 
  as.data.frame()

df2[1,2:n] <- stns
df2[2,2:n] <- reps
names(df) <- names(df2)
df <- df %>%
  mutate(across(all_of(names(df)), as.character))

df <- df2 %>%
  bind_rows(df)

names(df) <- rep("", ncol(df))

wide_data_stns <- df
```

```{r data_wide_stns}
head(wide_data_stns)
```

Note that if the if the observation data *only* has stations *or* replicates, then rearranging the data can be done in one step, as in the [previous example](data with species names in columns). But in this case, the station and replicate information are in separate rows. The restructuring process is a little more complicated.

Before we can use a pivot function, we need to combine the station ID and replicate ID for each column into a single value. In this example, each station ID and replicate ID are joined into a single character value, with an underscore to separate them *`_`*. The underscore will be used again after pivoting the table to identify where to split the combined station/replicate values back into separate values again.

If there are station names which already contain underscores, then another suitable character should be used when joining and splitting station/replicate IDs.

```{r combine_stn_rep}
sep_character <- "_"

# get the station IDs from row 1, excluding column 1 (this contains species names)
stations <- wide_data_stns[1, 2:ncol(wide_data_stns)]

# get the replicate IDs from row 2
replicates <- wide_data_stns[2, 2:ncol(wide_data_stns)]

# combine the station and replicate for each column using an underscore
station_replicate <- paste(stations, replicates, sep=sep_character)

# now we have extracted the station/replicate information, we can drop the 
# first two rows of the data frame
wide_data_stns <- wide_data_stns[3:nrow(wide_data_stns),]

# apply the station_replicates as column names
names(wide_data_stns) <- c("species", station_replicate)

head(wide_data_stns)
```

We can see that the column names now contain the combined station and replicate information. We are ready to transpose the data.

```{r pivot_stations}
# column 1 contains species names
# so, select all columns from 2 to be pivoted 

long_data <- wide_data_stns %>%
  pivot_longer(cols = 2:ncol(wide_data_stns), 
               names_to = "stn_rep",
               values_to = "count")

head(long_data)
```

Now we can split the *stn_rep* column into separate columns for *station* and *replicate*, We will use the underscore we introduced earlier to indentify where the split should occur:

```{r split_stations}

long_data <- long_data %>%
  separate_wider_delim(cols="stn_rep", 
                       delim = sep_character,
                       names=c("station", "replicate"))

head(long_data)
```

Now we are ready to calculate AMBI...

# Calculate AMBI

[Species groups](./about-AMBI.html#species-groups)

# Calculate M-AMBI

Calculate M-AMBI the multivariate AMBI index, based on the three separate species diversity metrics:

-   AMBI index `AMBI`.
-   Shannon Wiener diversity index `H'`
-   Species richness `S`.

All three indices are returned by the `AMBI()` function and therefore the results include all necessary input data for `MAMBI()`.
