---
title: "Interactive mode"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interactive mode}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

By default, the `AMBI()`function works in the same way as many (_most?_) other
functions you are familiar with: it is called in an _unsupervised_ manner. 
That is, the function is called and then, hopefully, returns a set of results 
without any additional interaction.

After checking the results, we can check if there are any species which were not 
recognized and, if necessary, call the function again, in an iterative process.

However, it is also possible to call the function in an _interactive_ mode. This
is similar to the way the original `AMBI`software functions, in that it is 
possible to interact with the function before the results are generated. Where
species names in the observation data were not recognized, you can manually 
assign the species to an `AMBI` species group.

```{r setup}
library(ambiR)
```

## Test data

We create a test data set with species names and counts

```{r data}
df <- data.frame(station = c("1","1","1","1","1","1",
                             "2","2","2","2","2","2"),
                 species = c("Acidostoma neglectum",
                             "Acrocirrus validus",
                             "Acteocina bullata",
                             "Austrohelice crassa",
                             "Capitella nonatoi",
                             "Watersipora subatra",
                             "Zanclea giancarloi",
                             "Zanclea sessilis",
                             "Zonorhynchus sp.",
                             "Acholoe squamosa",
                             "Aglaophenia kirchenpaueri",
                             "Tricellaria sp."),
                 count = c(2, 4, 5, 3, 7, 2, 4, 5, 3, 7, 2, 3))
```

## Instructions

The `AMBI()` function is called using the `interactive = TRUE` argument.

If any of the species names in the input data frame were not recognized as 
belonging to the `AMBI` species list *or* to a user-defined species list, if 
this was specified, then a message is displayed showing the number of 
unrecognized species, and giving instructions for how to proceed with assigning 
groups to each of these species.

The first species in the list of unrecognized names will be displayed. To proceed
you should type at the console prompt:

-   `1`,`2`,`3`,`4` or `5` - the species will be assigned to the selected
([AMBI species groups](./about-AMBI.html#species-groups))

-   `0` - assigning Group `0` means that the species will be inclued in the AMBI 
calculations and count towards the number of *Not assigned* species.

-   `Enter` - the species name will be treated as unrecognized and will *not* 
be included in `AMBI` calculations.

-   `s` - a list of recognized species names from the `AMBI` list will be 
displayed. These are species which appear close to the the unrecognized 
species when they are sorted in alphabetical order. 10 species names will be 
displayed. Typing `s` repeatedly will display consecutive names from the `AMBI`
list, 10 at a time.

-   `x` - the interactive species assignment process will be aborted. Any 
species groups assigned manually up to this point will be discarded and the 
AMBI calculations will continue using the default, unsupervised, method.

## Interactive AMBI calculation

```{r step1, eval=FALSE, include=TRUE}
AMBI(df, by = c("station"), quiet=F, interactive = T)
#> 
#> ── Assigning unrecognized species ──────────────────────────────────────────────
#> 
#> • 7 species names were not recognized. These will now be displayed, one at a
#> time.
#> 
#> • For each species, you can assign it to one of the five AMBI categories (I,
#> II, III, IV, V). Do this by entering an integer value from 1 to 5.
#> Alternatively, you can assign a value of 0. This indicates that the species
#> name is recognized but it is not possible to assign it to one of the five
#> categories.
#> 
#> • If Enter is pressed, without providing a value, no change will be made. The
#> species name in question will be treated as unrecognized.
#> 
#> • Enter s at the prompt to see a list of similar species names and their
#> corresponding AMBI categories.
#> 
#> • Enter x to abort the interactive species selection. Any entries made up to
#> that point will be discarded. All 7 species will be treated as unrecognized.
#> 
#> ── 1 Acholoe squamosa ──
#> 
#> ℹ enter an integer value from 0 to 5 or press Enter to skip. s - see similar
#>  names. x - abort interactive assignment.
```

Typing `s` will show species names which are close to the unrecognized species
when arranged in alphabetical order.

```{r step2, eval=FALSE, include=TRUE}
s

#> • Achelous ordwayi                                 (Group 1)
#> • Acidostoma hancocki                              (Group 2)
#> • Acidostoma neglectum                             (Group 0)
#> • Acidostoma nodiferum                             (Group 1)
#> • Acidostoma obesum                                (Group 1)
#> • Acidostoma sarsi                                 (Group 1)
#> • Acidostoma sp.                                   (Group 1)
#> • Acila castrensis                                 (Group 2)
#> • Acila insignis                                   (Group 2)
#> • Aclis gulsonae                                   (Group 1)
#> ℹ enter an integer value from 0 to 5 or press Enter to skip. s - see similar
#>  names. x - abort interactive assignment.
```

There were no similar species names which might give us a clue about this 
species. In this case, we are satisfied that the name is a valid species 
name[^1] but we are unsure which group the species should be assigned to, so 
we assign group `0`.

[^1]: <https://www.marinespecies.org/aphia.php?p=taxdetails&id=146474>

Assigning group `0` means that the species will be included in the count of
`Not assigned` species when calculating the `AMBI` index. If we typed `Enter`
the species would be ignored and not included in the `AMBI` calculations.

A message is displayed confirming our selection (`Not assigned`) and then we
are asked to consider the next unrecognized species name,
*Aglaophenia kirchenpaueri*.

```{r step3, eval=FALSE, include=TRUE}
0

#> ✔ Acholoe squamosa - Not assigned
#> 
#> ── 2 Aglaophenia kirchenpaueri ──
#>
#> ℹ enter an integer value from 0 to 5 or press Enter to skip. s - see similar
#>  names. x - abort interactive assignment.
```

We type `s` to see similar names:

```{r step4, eval=FALSE, include=TRUE}
s

#> • Aglaophamus verrilli                             (Group 2)
#> • Aglaophenia acacia                               (Group 0)
#> • Aglaophenia pluma                                (Group 0)
#> • Agnezia septentrionalis                          (Group 2)
#> • Aidanosagitta crassa                             (Group 1)
#> • Aiptasia diaphana                                (Group 3)
#> • Aiptasia mutabilis                               (Group 3)
#> • Akanthophoreus gracilis                          (Group 1)
#> • Akera bullata                                    (Group 1)
#> • Akera sp.                                        (Group 1)
#> ℹ enter an integer value from 0 to 5 or press Enter to skip. s - see similar
#>  names. x - abort interactive assignment.
```

In this case we can see two species belonging to the same genus, _Aglaophenia_.
They are both assigned to group `0` (`Not assigned`) so we could make the 
assumption that our unrecognized species should be assigned to the same group.

```{r step5, eval=FALSE, include=TRUE}
0

#> ✔ Aglaophenia kirchenpaueri - Not assigned
#>
#> ── 3 Tricellaria sp. ──
#>
#> ℹ enter an integer value from 0 to 5 or press Enter to skip. s - see similar
#>  names. x - abort interactive assignment.
```

This process is repeated for each unrecognized species name.

## Results

The `AMBI()` function returns a list of three dataframes, including `matched`. 
This dataframe shows which groups species have been assigned to and after 
interactive species matching, it will include the column `source`.
An _'I'_ in this column indicates that the group was assigned interactively.

```{r matched, eval=FALSE, include=TRUE}
$matched
   station                   species           species_matched count group source RA
1        1      Acidostoma neglectum      Acidostoma neglectum     2     0   <NA>  0
2        1        Acrocirrus validus        Acrocirrus validus     4     4   <NA>  0
3        1         Acteocina bullata         Acteocina bullata     5     1   <NA>  0
4        1       Austrohelice crassa       Austrohelice crassa     3     5   <NA>  0
5        1         Capitella nonatoi         Capitella nonatoi     7     5   <NA>  0
6        1       Watersipora subatra       Watersipora subatra     2     0      I NA
7        2        Zanclea giancarloi        Zanclea giancarloi     4     0      I NA
8        2          Zanclea sessilis          Zanclea sessilis     5     0      I NA
9        2          Zonorhynchus sp.          Zonorhynchus sp.     3     0      I NA
10       2          Acholoe squamosa          Acholoe squamosa     7     0      I NA
11       2 Aglaophenia kirchenpaueri Aglaophenia kirchenpaueri     2     0      I NA
12       2           Tricellaria sp.           Tricellaria sp.     3     3      I NA
```

## Aborting interactive mode

If, at any point in the interactive species assignment process we type `x` 
then the interactive process will be aborted. You will see the following message

```{r step6, eval=FALSE, include=TRUE}
x

#> ! interactive species assignment cancelled
```

Any species groups assigned manually up to this point will be discarded and 
the AMBI calculations will proceed using the default, unsupervised, method.
